---
title: "Analysis of airline on-time performance data for the year 2008"
author: "Liu_Xiaotian"
date: "2024-05-31"
output: html_document
---

## 1. Introduction

This project analyzes airline on-time performance data for the year 2008. We aim to identify the best times of day, days of the week, and times of the year to minimize delays, uncover the primary factors contributing to delays and cancellations, and determine which flights are most frequently affected.

## 2. Data Loading

The data used in this analysis was processed using Apache Pig scripts and saved in CSV format. Below, we load this data into R for analysis.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
setwd('/Users/liuxiaotian/Documents/P136558_LiuXiaotian_DataManagementAssignment2')

avg_delays_by_day <- read.csv("avg_delays_by_day.csv", header = FALSE)
avg_delays_by_time <- read.csv("avg_delays_by_time.csv", header = FALSE)
cancellations_by_code <- read.csv("cancellations_by_code.csv", header = FALSE)
flight_problems <- read.csv("flight_problems.csv", header = FALSE)
names(avg_delays_by_day) <- c("day_of_week", "avg_dep_delay")
names(avg_delays_by_time) <- c("dep_time", "avg_dep_delay")
names(cancellations_by_code) <- c("cancellation_code", "total_cancellations")
names(flight_problems) <- c("Carrier", "Year", "Flights", "Avg_Delay")
flight_problems$Carrier <- gsub("[()]", "", flight_problems$Carrier)
flight_problems$Year <- gsub("[()]", "", flight_problems$Year)
flight_problems$Year <- as.numeric(flight_problems$Year)
flight_problems$Avg_Delay <- as.numeric(flight_problems$Avg_Delay)
```

## 3. Analysis

### 3.1 Average Delays by Day of the Week

This section explores which days of the week experience the least and most delays, helping travelers choose the best day to fly.

```{r}
ggplot(avg_delays_by_day, aes(x = day_of_week, y = avg_dep_delay)) +
  geom_col(fill = "steelblue") +
  labs(title = "Average Departure Delays by Day of Week", x = "Day of the Week", y = "Average Delay (minutes)")
```

This chart shows the average departure delays by day of the week. Days are represented as numbers, with 1 being Monday and 7 Sunday. The chart highlights that mid-week, specifically Wednesday (Day 3) and Friday (Day 5), experiences higher average delays compared to other days. This could be due to higher travel volumes mid-week and end of the workweek.

### 3.2 Average Delays by Departure Time

We analyze how departure time affects flight delays, potentially aiding in choosing optimal flight times.

```{r}
ggplot(avg_delays_by_time, aes(x = dep_time, y = avg_dep_delay)) +
  geom_line(color = "red") +
  labs(title = "Average Departure Delays by Departure Time", x = "Scheduled Departure Time", y = "Average Delay (minutes)")
```

This chart shows the average departure delays across different scheduled departure times, plotted throughout the day from 0 to 2400 hours. Significant spikes can be observed at specific times, likely during peak hours such as early morning and late evening. These spikes may indicate congestion and operational delays at airports during these times.

### 3.3 Flight Cancellations by Reason

This analysis determines the most common reasons for flight cancellations.

```{r}
ggplot(cancellations_by_code, aes(x = cancellation_code, y = total_cancellations, fill = code)) +
  geom_bar(stat = "identity", fill = 'pink') +
  labs(title = "Flight Cancellations by Reason", x = "Cancellation Code", y = "Total Cancellations")
```

This graph depicts the distribution of flight cancellations by reason, labeled by codes. The concentration of cancellations in lower code numbers suggests that reasons classified under these codes are more frequent. This distribution could help identify the most common issues leading to cancellations.

### 3.4 Most Frequently Delayed and Cancelled Flights

Identifying specific flights that are most problematic in terms of delays and cancellations.

```{r}
ggplot(flight_problems, aes(x = Year, y = Avg_Delay, group = Carrier, color = Carrier)) +
  geom_line() +
  geom_point() +
  labs(title = "Average Delay by Year for Each Carrier", x = "Year", y = "Average Delay (minutes)")
```

This colorful plot illustrates the average delay experienced by each carrier over the years. It shows significant variability between different carriers and over different years. Some carriers consistently show higher delays, which could reflect operational challenges or route complexities specific to those carriers.

## 4. Conclusion

The analysis of flight data highlights significant variability in departure delays related to specific times of the day and week, as well as operational challenges faced by different carriers. Understanding these patterns can aid in improving scheduling and operational efficiency in the aviation industry.